The Ranked Nearest Neighbour Interchange space

### 2019-08-21 paper.tex
restructure: diameter and radius section

\newcommand{\np}{\mathcal{NP}}
\newcommand{\parent}{\mathrm{parent}}
\newcommand{\mrca}{\mathrm{mrca}}
\newcommand{\rank}{\mathrm{rank}}
\newcommand{\nni}{\mathrm{NNI}}
\newcommand{\rnni}{\mathrm{RNNI}}
\newcommand{\rnniu}{\mathrm{RNNIu}}
\newcommand{\tbr}{\mathrm{TBR}}
\newcommand{\spr}{\mathrm{SPR}}
\newcommand{\csort}{\textsc{Caterpillar Sort}}
\newcommand{\findpath}{\textsc{FindPath}}

@@

\subsection{Diameter}
\label{section:diameter}

With the results of Section~\ref{section:caterpillar_convex} we can easily provide the exact diameter of the $\rnni$ graph and hence improve the bounds for the diameter given in \autocite{Gavryushkin2018-ol}.
Moreover, we can conclude from Lemma~\ref{lemma:caterpillar_diameter} that the maximum distance computed by algorithm $\findpath$, introduced in Section~\ref{section:geometry}, equals the diameter of $\rnni$.
This suggests that this algorithm computes reasonable estimates of distances.
For further details on $\findpath$ we refer the reader to Section~\ref{section:computation}.

@@

@@

Recall that the Section~\ref{section:diameter} revealed that the maximum distance computed by $\findpath$ for an arbitrary pair of trees equals the diameter of $\rnni$.
Thus, if the distance between two trees equals the diameter, $\findpath$ computes the correct distance.
Following this observation, we claim following conjecture:

@@

Besides the diameter, the \emph{radius} of a graph $G$, which is defined by $rad(G) =  \min\limits_u \max\limits_v d(u,v)$ is a further interesting property of a graph.
We will use Algorithm~\ref{alg:max_dist_tree} to show that in $\rnni$ the radius equals the diameter.

%TODO
%Explain the following algorithm, not only Pseudo-Code!

\begin{algorithm}[H]
\caption{MAX\_DISTANCE\_TREE($T$)}
\label{alg:max_dist_tree}
\begin{algorithmic}[1]
	\STATE $R$ only contains cherry taxa $a, b$ of $T$ and has internal node with rank $n-1$
    \STATE In list $L$ all taxa $\{1,\ldots,n\}\setminus\{a,b\}$ are sorted such that the ranks of their parents in $T$ do not decrease
	\FOR {taxon $a$ in $L$}
		\STATE Add an internal node as parent of $a$ on an edge in $R$ such that the new internal node has rank one less than the previously lowest internal node
	\ENDFOR
	\RETURN $R$
\end{algorithmic}
\end{algorithm}

\begin{lemma}
    Let $T$ be a caterpillar tree.
    A tree $R$ that is computed by applying Algorithm~\ref{alg:max_dist_tree} on $R$ has distance $d(T,R) = \Delta(\rnni)$ to $T$.
    \label{lemma:max_dist_caterpillar}
\end{lemma}

    Induction basis: $n = 3$\\
    Algorithm~\ref{alg:max_dist_tree} computes either $R = ((1,3),2)$ or $R'= ((2,3),1)$ when applied to $T = ((1,2),3)$.
    Since it is $d(T,R) = d(T,R') = 1 = \Delta(\rnni_3)$, the lemma is true for trees on $n=3$ taxa.

    Induction step: $n \to n+1$\\
    Induction hypothesis: For a caterpillar tree $T$ on $n$ taxa Algorithm~\ref{alg:max_dist_tree} computes a tree $R$ with distance $d(T,R) = \frac{(n-1)(n-2)}{2}$.\\
    Let us now  assume that $T$ is a tree on $n+1$ taxa.
    In the list $L$ produced in Algorithm~\ref{alg:max_dist_tree}, when applied to $T$, taxon $n+1$ is the last one.
    Hence $n+1$ is added to $R$ in the last step and therefore, it is in the lowest cherry of $R$ and has parent with rank $1$.
    Because the parent of $n+1$ has rank $n$ in $T$, we can conclude from Lemma~\ref{lemma:distance_delete_taxon} that for $T$ and $R$ restricted to taxa $1,\ldots,n$ it holds: $d(T|_n,R|_n) \leq d(T,R) - (n-1)$.
    It is easy to see that $R|_n$ can be received from $T|_n$ by applying Algorithm~\ref{alg:max_dist_tree}.
    So we can follow from the induction hypothesis and Lemma~\ref{lemma:distance_delete_taxon} that it is $d(T,R) \geq d(T|_n,R|_n) + (n-1) = \frac{(n-1)(n-2)}{2} = \frac{n(n-1)}{2}$, which concludes the proof of the lemma.
\end{proof}

In Algorithm~\ref{alg:max_dist_tree} it is not uniquely determined, where a new taxon is added to the already existing tree.
Instead, the attachment branch can be randomly chosen with the only restriction being that the rank of the new node needs to be below the ranks of already existing internal nodes.
This is the reason why one can receive every tree topology on $n$ taxa when applying Algorithm~\ref{alg:max_dist_tree} onto a caterpillar tree.
And because permuting the labels of the trees does not change the distance, there is for every tree $R$ on $n$ taxa a caterpillar tree $T$ with distance $d(T,R) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
This gives us Corollary~\ref{corollary:radius}.

>>>>

\newcommand{\np}{\mathcal{NP}}
\newcommand{\parent}{\mathrm{parent}}
\newcommand{\mrca}{\mathrm{mrca}}
\newcommand{\rank}{\mathrm{rank}}
\newcommand{\nni}{\mathrm{NNI}}
\newcommand{\rnni}{\mathrm{RNNI}}
\newcommand{\rnniu}{\mathrm{RNNIu}}
\newcommand{\tbr}{\mathrm{TBR}}
\newcommand{\spr}{\mathrm{SPR}}
\newcommand{\csort}{\textsc{Caterpillar Sort}}
\newcommand{\findpath}{\textsc{FindPath}}
\newcommand{\mdtree}{\textsc{MDTree}}

@@

\subsection{Diameter and radius}
\label{section:diameter}

With the results of Section~\ref{section:caterpillar_convex} we can easily provide the exact diameter of the $\rnni$ graph and hence improve the bounds for the diameter given in \autocite{Gavryushkin2018-ol}.
Furthermore, we are able to show that the \emph{radius} of the $\rnni$ graph, which is defined as $rad(G) =  \min\limits_u \max\limits_v d(u,v)$ for a graph $G$, equals its diameter.
Let us first consider the diameter of the $\rnni$ graph.
With the results of Section~\ref{section:caterpillar_convex} is not hard to see that is $\Delta(\rnni) = \frac{(n-1)(n-2))}{2}$ as it is stated in Corollary~\ref{corollary:diameter}.

@@

We now proceed to show that the radius of the $\rnni$ graph equals its diameter.
For this aim an algorithm for constructing a tree $R$ with distance $\Delta(\rnni)$ from a given caterpillar tree $T$ will be used.
This algorithm will be called $\mdtree$ (Maximum Distance Tree).
We will prove that a tree $R = \mdtree(T)$ has distance $d(T,R) = \Delta(\rnni)$ to $T$ if $T$ is a caterpillar tree, after introducing the algorithm.

$\mdtree$ works as follows:
At first, all taxa of $T$ are sorted in a list $L$ such that the ranks of their parents are non-decreasing.
If the input is a caterpillar tree, it follows that $L[0]$ and $L[1]$ are the cherry taxa of $T$ and $\parent(L[i])$ has rank $i$ in $T$ for all $i = 2, \ldots, n-1$.
Now $R$ is built inductively as follows:
Initially, $R$ only consists of a cherry with taxa $L[0]$ and $L[1]$ such that the parent of these taxa has rank $n-1$.
All following taxa in $L$ are added to $R$ sequentially, according to their order in $L$, by creating a new internal node at one of the existing branches in $R$ such that this new node has rank one less than the previously lowest internal node in $R$.
When this is done for all taxa in $L$, the resulting tree $R$ is an $\rnni$ tree.

It is important to note that this algorithm can be applied to any input tree $T$, which does not need to be a caterpillar tree.
However, the results provided in this section are only true for caterpillar trees.
The pseudocode for $\mdtree$ can be found in Section~\ref{section:computation}.
\todo{figure? + do we need the pseudocode?}

\begin{lemma}
    Let $T$ be a caterpillar tree.
    A tree $R = \mdtree(T)$ has distance $d(T,R) = \Delta(\rnni)$ to $T$.
    \label{lemma:max_dist_caterpillar}
\end{lemma}

    Induction basis: $n = 3$\\
    $\mdtree(T)$ results either in $R = ((1,3),2)$ or $R'= ((2,3),1)$ for $T = ((1,2),3)$.
    Since it is $d(T,R) = d(T,R') = 1 = \Delta(\rnni_3)$, the lemma is true for trees on $n=3$ taxa.

    Induction step: $n \to n+1$\\
    Induction hypothesis: For a caterpillar tree $T$ on $n$ taxa $\mdtree$ computes a tree $R$ with distance $d(T,R) = \frac{(n-1)(n-2)}{2}$.\\
    Let us now  assume that $T$ is a tree on $n+1$ taxa.
    In the list $L$ produced in $\mdtree(T)$ the taxon $n+1$ is the last one, $L[n] = n+1$.
    Hence $n+1$ is added to $R$ in the last step and therefore, it is in the lowest cherry of $R$ and has parent with rank $1$.
    Because the parent of $n+1$ has rank $n$ in $T$, we can conclude from Lemma~\ref{lemma:distance_delete_taxon} that for $T$ and $R$ restricted to taxa $1,\ldots,n$ it holds: $d(T|_n,R|_n) \leq d(T,R) - (n-1)$.
    It is easy to see that $R|_n$ can be received from $T|_n$ by applying $\mdtree$.
    So we can follow from the induction hypothesis and Lemma~\ref{lemma:distance_delete_taxon} that it is $d(T,R) \geq d(T|_n,R|_n) + (n-1) = \frac{(n-1)(n-2)}{2} = \frac{n(n-1)}{2}$, which concludes the proof of the lemma.
\end{proof}

In $\mdtree$ it is not uniquely determined, where a new taxon is added to the already existing tree.
Instead, the attachment branch can be randomly chosen with the only restriction being that the rank of the new node needs to be below the ranks of already existing internal nodes.
Therefore the output tree $R$ of $\mdtree$ is not unique.
Every tree topology on $n$ taxa can be resulting from $\mdtree$.
And because permuting labels of trees does not change the distance between them, $\mdtree$ shows that there is for every tree $R$ on $n$ taxa a caterpillar tree $T$ with distance $d(T,R) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
This gives us Corollary~\ref{corollary:radius}.

@@

Recall that Section~\ref{section:diameter} revealed that the maximum distance computed by $\findpath$ for an arbitrary pair of trees equals the diameter of $\rnni$.
Thus, if the distance between two trees equals the diameter, $\findpath$ computes the correct distance.
Following this observation, we claim Conjecture~\ref{conjecture:findpath_maxdist_correct}.
Notice that one direction of the conjecture is already known to be true, it only remains open whether the distance of two trees is $\Delta(\rnni)$ indeed, if $\findpath$ computes a path of this length.

@@

In order to find the radius of the $\rnni$ graph, the algorithm $\mdtree$ has been introduced in Section~\ref{section:diameter}.
It computes a tree $R$ with distance $\Delta(\rnni)$ from a given caterpillar tree, as it has been proven in Lemma~\ref{lemma:max_dist_caterpillar}.
Algorithm~\ref{alg:max_dist_tree} provides the pseudocode for this algorithm.
It is important to note that $\mdtree$ can not only be applied onto caterpillar trees, as it is done in section~\ref{section:diameter}, but to any arbitrary tree.
However, it does not necessarily produce a tree with maximum distance from a given tree.

\begin{algorithm}[H]
\caption{$\mdtree(T)$}
\label{alg:max_dist_tree}
\begin{algorithmic}[1]
	\STATE $R$ only contains cherry taxa $a, b$ of $T$ and has internal node with rank $n-1$
    \STATE In list $L$ all taxa $\{1,\ldots,n\}\setminus\{a,b\}$ are sorted such that the ranks of their parents in $T$ do not decrease
	\FOR {taxon $a$ in $L$}
		\STATE Add an internal node as parent of $a$ on an edge in $R$ such that the new internal node has rank one less than the previously lowest internal node
	\ENDFOR
	\RETURN $R$
\end{algorithmic}
\end{algorithm}
